Static contract checking for Haskell

### 2011-08-17 paper.tex
New version with

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\cfc}{\texttt{CF}}

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.
There are two different symbols for implication : $\to$ and
$\implies$. Altough they share the same semantics, their operational behaviour differs, as explained in \ref{guarded-pat}.

\begin{figure}
  \label{fol-stx}
  \centering
  \begin{array}{rcl}
    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
    && \mid \bad \mid \unr & \\
    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
    && \mid \phi \land \phi \mid \phi \implies \phi \mid \phi \to \phi \mid \phi \leftrightarrow \phi & \\
    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
  \end{array}
  \caption{First-order logic syntax}
\end{figure}

@@

\begin{figure}
  \label{dtrans-reg}
  $$\dtrans{f(x_1,\dots,x_n) = e} = Y = f(X_1,\dots,X_n) \implies Y = \dtrans{e}$$
  \caption{$\dtrans{}$ -- Regular definitions}
\end{figure}

\begin{figure*}
  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_i(\overline{z})] = e_i } = \bigwedge_{1 \leq i \leq n} Y = f(K_i(x_1,\dots,x_{K_i})) \implies Y = \etrans{e_i} $$
  $$ \cup f(X_1,\dots,X_n) = Y \implies Y = \unr \lor \bigvee_{1 \leq i \leq n} \etrans{e} = K_i(sel_{1,K_i}(\etrans{e}),\dots,sel_{n_K,K_i}(\etrans{e})) \lor \etrans{e} = \bad $$
  \caption{$\dtrans{}$ -- Case definitions}
\end{figure*}

@@

\begin{itemize}
\item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i)$
  on its $k$-th component.
\item $(S_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(S_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\begin{figure*}
  \label{ttrans}
  $$ S_1 := \bigwedge_{1 \leq i \leq n} sel_{i,K_j}(K_j(X_1,\dots,X_n)) = Y \implies Y = X_i  \mid 1 \leq j \leq n$$
  $$ S_2 := \forall x_1,\dots,x_{n_i}~\forall y_1,\dots,y_{n_j} . K_i(x_1,\dots,x_{n_i}) \neq K_j(y_1,\dots,y_{n_j}) \mid 1 \leq i < j \leq n$$
  $$ S_3 := \forall x_1,\dots,x_{n_i} . (\cf{x_1} \land \dots \land \cf{x_{n_i}} \leftrightarrow \cf{K_i(x_1,\dots,x_{n_i})}) \mid 1 \leq i \leq n$$
  $$ S_4 := \forall x_1,\dots,x_n.~K_i(x_1,\dots,x_n) \neq \unr \mid 1 \leq i \leq n$$
  \caption{$\ttrans{}$ -- Data type translation}
\end{figure*}

\subsection{$\strans{}$ -- Contracts}
We give in \ref{strans} the translation of contract satisfaction.

\begin{figure*}
$$ \strans{f \in c} = \top \to f \in c \lor \bot $$

Given $r$ a fresh predicate and $a$ a fresh variable,
\begin{eqnarray*}
  p \to f \in \{x \mid b(x) \} \lor q &=& p \land f=X \implies b(X)=true \lor b(X) = \unr \lor q  \\
  p \to f \in x:c_1 \to c_2(x) \lor q &=&  p \to f(X) \in c_2(X) \lor r(X), r(X) \to X \not \in c_1 \lor q \\
  p \to f \in \cfc \lor q &=& p \land \forall x_1,\dots,x_n. \bigwedge_i \cf{x_i} \to \cf{f(x_1,\dots,x_n)} \lor q 
\end{eqnarray*}

\begin{eqnarray*}
  p \to f \not \in \{x \mid b(x) \} \lor q &=&  p \land f=X \implies b(X)=false \lor b(X) = \bad \lor q  \\
  p \to f \not \in x:c_1 \to c_2(x) \lor q &=&  p \to a \in c_1 \lor q, p \to f(a) \not \in c_2(a) \lor q \\
  p \to f \not \in \cfc \lor q &=&  p \land \forall x_1,\dots,x_n. \lnot \bigwedge_i \cf{x_i} \to \cf{f(x_1,\dots,x_n)} \lor q 
\end{eqnarray*}
\caption{$\strans{}$ -- Contract translation}
\label{strans}
\end{figure*}

@@

\section{Guarded patterns}
It is our first encounter with the guarded-pattern operator
($\implies$). 
It syntax is:
$$ s_1 = t_1 \land \dots \land s_k = t_k \implies v_1 = w_1 \lor \dots \lor v_n = w_n $$

$s_i,t_i,v_i,w_i$ are terms, and all the variables on the right-hand
side should appear somewhere on the left-hand side.

The semantics are: if during the proof by the theorem prover, we have
equations that match the LHS than the theorem prover should strongly
consider the equations on the RHS. Those semantics are sound (we can
interpret $\implies$ as normal implication) but if modifies the
operational behabviour of Equinox, because it tells it very finely
when to use equations and how to instantiate them.

\section{Higher-orderness}
\label{ho}
Our current translation only considers fully applied functions and
first-order functions. For example, we cannot translate any contract
for $map$ because it would involve quantifying over a function, a
thing that is not first-order.

There is a possible workaround, which involves the ``app'' function in
first-order logic. Assume we have a function $f$ that is not fully
applied somewhere in a module. We create the term $f\_ptr$ which
relates to $f$ by the equations given in \ref{ho-fig}

This way, we can emulate quantification over function by
quantification on their $ptr$ counterpart.

\begin{figure*}
\label{ho-fig}
$$\etrans{e_1~e_2} = app(e_1,e_2)$$
$$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$
$$ \cf{f\_ptr} \iff \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \implies \cf{f(x_1,\dots,x_n)}$$
$$\forall f\_ptr,x~\cf{f\_ptr} \land \cf{x} \implies \cf{app(f\_ptr,x)}$$
\caption{Encoding of higher-orderness}
\end{figure*}

>>>>

\newcommand{\tot}{\leftrightarrow}

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
\renewcommand{\min}[1]{\mbox{$\$$min}(#1)}

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.

\begin{figure}
  \label{fol-stx}
  \centering
  \begin{array}{rcl}
    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
    && \mid \bad \mid \unr & \\
    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi & \\
    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
  \end{array}
  \caption{First-order logic syntax}
\end{figure}

@@

\begin{figure}
  \label{dtrans-reg}
  $$\dtrans{f(\vec{x}) = e}$$
  $$ \forall \vec{x},y.~ \min{y} \land y = f(\vec{x}) \to y = \etrans{e}$$
  \caption{$\dtrans{}$ -- Regular definitions}
\end{figure}

\begin{figure*}
  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
  \begin{center}
    \begin{array}{rccccccl}
      \forall \vec{a},y,\vec{x}_{1 \leq i \leq n}.~&\min{y} &\land& \etrans{e} = K_1(\vec{x_1}) &\land& f~\vec{a} = y &\to& y = \etrans{e_1}\\
                                               &&&&&& \cdots & \\
                                               &\min{y} &\land& \etrans{e} = K_n(\vec{x_n}) &\land& f~\vec{a} = y &\to& y = \etrans{e_n}\\
                                               &\min{y} &\land& \etrans{e} = \bad          &\land& f~\vec{a} = y &\to& y = \bad\\
                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{a} \land (a=\bad \lor \bigvee_i a = \weak{K_i(\vec{x})} \lor y = \unr)
    \end{array}
    \end{center}
  \caption{$\dtrans{}$ -- Case definitions}
\end{figure*}

@@

\begin{itemize}
\item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
  on its $k$-th component.
\item $(S_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(S_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\begin{figure*}
  \label{ttrans}
  $$ S_1 := \forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = \weak{sel_{j,K_i}(a)}  \mid 1 \leq i \leq n$$
  $$ S_2 := \forall \vec{x},\vec{y},a,b.~(\min{a} \land \min{b} \land K_i(\vec{x}) = a \land K_j(\vec{y}) = b) \to a \neq b \mid 1 \leq i < j \leq n$$
  $$ S_3 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to (\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})} \mid 1 \leq i \leq n$$
  $$ S_4 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad \mid 1 \leq i \leq n$$
  \caption{$\ttrans{}$ -- Data type translation}
\end{figure*}

\subsection{$\strans{}$ -- Contracts}
We give in \ref{strans} the translation of contract satisfaction.
$true$ refers to the translation to a term of the data constructor
\texttt{True} in $\hprime$, not to the actual true value.

\begin{figure*}
\begin{eqnarray*}
  \strans{e ::: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
  \strans{e ::: x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x :/: c_1} \lor  \strans{\etrans{e(x)} ::: c_2(x)})\\
  \strans{e ::: \cfc} &=& \cf{e}
\end{eqnarray*}

\begin{eqnarray*}
  \strans{e :/: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} \neq true \land \etrans{b(e)} \land \unr)\\
  \strans{e :/: x:c_1 \to c_2(x)} &=& \exists x.~\min{\etrans{f(x)}} \to (\strans{x ::: c_1} \land  \strans{\etrans{e(x)} :/: c_2(x)})\\
  \strans{e :/: \cfc} &=& ~\cf{e}
\end{eqnarray*}

$$ \strans{e ::: c_1 && c_2} = \strans{e ::: c_1} \land \strans{e ::: c_2} $$
$$ \strans{e ::: c_1 || c_2} = \strans{e ::: c_1} \lor \strans{e ::: c_2} $$
$$ \strans{(a,b) ::: (c_1,c_2)}   = \strans{a ::: c_1} && \strans{b ::: c_2} $$

\caption{$\strans{}$ -- Contract translation}
\label{strans}
\end{figure*}

@@

\section{Higher-orderness}
\label{ho}
Our current translation only considers fully applied functions and
first-order functions. For example, so far, we cannot give any
contract to $map$ because it would involve quantifying over a
function, a thing that is not first-order.

There is a possible workaround, which involves the ``app'' function
defined in our first-order logic. Assume we have a function $f$ that
is not fully applied somewhere in a module. We create the term
$f\_ptr$ which relates to $f$ by the equations given in \ref{ho-fig}

This way, we can emulate quantification over function by quantifying
on their $ptr$ counterpart.

\begin{figure*}
\label{ho-fig}
$$\etrans{e_1~e_2} = app(e_1,e_2)$$
$$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$
$$ \cf{f\_ptr} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{f(x_1,\dots,x_n)}$$
$$\forall f\_ptr,x.~\cf{f\_ptr} \land \cf{x} \to \cf{app(f\_ptr,x)}$$
\caption{Encoding of higher-orderness}
\end{figure*}
